Think deep for reef resilience

### 2025-09-13 paper/paper_draft.qmd
Updated text describing time series clustering and RF

Reefs were clustered into three qualitative categories based on their temporal behaviour
indicated by coral cover over time. The three categories were Low, Medium, and High,
indicating their relative ability to maintain coral cover and thus their structural
resilience. The Complexity-Invariant Distance (CID) between each time series was calculated
to identify similarities in temporal behaviour [@batista2014]. The CID applies a correction
factor to the Euclidean distance between pairs of time series to obtain a measure of their
similarity. The adopted approach ensures reefs are clustered based on their temporal
dynamics as opposed to, for example, naively clustering by their mean coral cover.
Clustering similar time series is achieved through the use of k-medoids clustering [cf.
@steinmann2020]. Further details on the time series clustering approach are found in the
supplementary materials.

The time series clustering was applied over the time frame between 2025 and 2075. Spatially,
the clustering analyses were conducted for each GBR management area and at the scale of each
reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were also performed at the
whole-of-GBR scale but were found to lead to identical conclusions about characteristics
influencing cluster cover compared to management area and bioregion scale analyses and so
are not shown here. To compare time series clustering results across GCMs, management area
scale clustering results are used to preserve coarse scale latitudinal differences between
regions, while maintaining ease of comparison and visualisation. These comparisons include
quantification of differences between low and high cover clusters across management regions
under each GCM using the CID between cluster median time series. To assess how different
reef characteristics influence time series clustering at a finer spatial scale, bioregion
results are used to capture patterns occurring within individual groups of reefs. Results of
all analyses at all clustering scales can be found at Grier and Iwanaga [-@grier2025] for
those interested.

Reefs in each time series cluster were compared to identify characteristics that differ
between Low, Medium and High clusters. These characteristics include reef median depth, mean
time series DHW, weighted incoming connectivity, outgoing connection strength, ratio of
source-to-sink larval connectivity, larval retention probability and reef carrying capacity
(coral habitable area).

@@

Partial Dependence Analysis was conducted to understand the relationship between features
and bioregion-scale time series cluster predictions using a trained Random Forest (RF)
classifier [@friedman2001]. The analysis utilised the reef characteristics of interest as
the predictors, including the median reef depth, mean DHW for the given GCM, Log10 -
weighted incoming connectivity, Log10 - weighted outgoing connectivity, and Log10 -
outgoing-to-incoming connectivity ratio. The full dataset consisted of the time series
cluster assignment of each of the 2429 reefs under the five GCMs explored, for a total of
12,145 samples. The dataset was randomly split into training (60%) and testing (40%) subsets
to ensure model generalisability. The RF classifier achieved 99% accuracy on the training
set and 62% on the test set.

While 99% accuracy suggests the classifier is overfit to the training data, the 66% test
accuracy indicates a meaningful relationship between modelled reef characteristics and their
assigned clusters has been captured. A 3-fold cross-validation was also conducted, arriving
at a similar 67% accuracy rate. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier is reliable.
Furthermore, F1 performance scores for Low, Medium, and High clusters were determined to be
0.7, 0.5 and 0.65 respectively, indicating balanced performance across the cluster
categories. Lower performance with the Medium cluster reflects the inherent difficulty of
classifying transitional states in natural systems.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual predictors were systematically (one-at-a-time) varied across a grid of
50 equally spaced points spanning the predictor's range, while all other features were left
at their modelled values. The procedure creates (1) a modified test dataset, where the
target predictor was set to the grid value for all samples, (2) probabilistic predictions
for all three cluster classes using the trained RF classifier, and (3) the mean predicted
probability for each class across all test observations. The median was also assessed (not
shown) but was found not to substantially influence the interpretation of results. The
process isolates the marginal effect of each predictor by averaging over the natural
distribution of all other representative reef characteristics in the test dataset.
Additionally, to assess possible interaction between reef depth and incoming connectivity,
two-way partial dependence analysis was performed on grids along the range of these two
features.

>>>>

Reefs were clustered based on the temporal behaviour of their median coral cover time
series. Put simply, reefs that have similar trends and trajectories are more likely to be
clustered together. Reefs are clustered into Low, Medium, or High, indicating their relative
ability to maintain coral cover over time. Reefs in each time series cluster were compared
to identify characteristics that differ between Low, Medium and High clusters. These
characteristics include reef median depth, mean time series DHW, weighted incoming
connectivity, outgoing connection strength, ratio of source-to-sink larval connectivity,
larval retention probability and reef carrying capacity (coral habitable area).

The Complexity-Invariant Distance (CID) between each time series was calculated to identify
similarities in temporal behaviour [@batista2014]. The CID applies a correction factor to
the Euclidean distance between pairs of time series to obtain a measure of their similarity.
The adopted approach ensures reefs are clustered based on their temporal dynamics as opposed
to, for example, naively clustering by their mean coral cover. Time series clustering is
achieved through the use of k-medoids clustering [cf. @steinmann2020]. Further details on
the time series clustering approach are found in the supplementary materials.

The time series clustering was applied over the time frame between 2025 and 2075. The
clustering was conducted for three spatial scales: (1) the whole of GBR; (2) each GBR
management area; and (3) for each reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). The
analyses was conducted at the different spatial scales as, conceptually, the temporal
behaviour constituting "low", "medium" or "high" coral cover may differ at the different
spatial scales. As it was, analyses at the whole-of-GBR scale were found to lead to
identical conclusions regarding reef characteristics which influence coral cover and so for
brevity are not shown here.

To compare time series clustering results across GCMs, management area scale clustering
results are used to preserve coarse scale latitudinal differences between regions, while
maintaining ease of comparison and visualisation. These comparisons include quantification
of differences between low and high cover clusters across management regions under each GCM
using the CID between cluster median time series. To assess how different reef
characteristics influence time series clustering at a finer spatial scale, bioregion results
are used to capture patterns occurring within individual groups of reefs. Results of all
analyses at all spatial scales can be found at Grier and Iwanaga [-@grier2025] for those
interested.

@@

Partial Dependence Analysis was conducted to understand the relationship between features
and bioregion-scale time series cluster predictions using a trained Random Forest (RF)
classifier [@friedman2001]. The analysis utilised the reef characteristics of interest as
the predictors, including the median reef depth, mean DHW for the given GCM, log10-weighted
incoming connectivity, log10-weighted outgoing connectivity, and log10-outgoing-to-incoming
connectivity ratio. The full dataset consisted of the time series cluster assignment of each
of the 2429 reefs under the five GCMs explored, for a total of 12,145 samples. The dataset
was randomly split into training (60%) and testing (40%) subsets to ensure model
generalisability. The RF classifier achieved 79% accuracy on the training set and 59% on the
test set.

A 3-fold cross-validation was also conducted, arriving at a similar 57% accuracy rate.
Furthermore, F1 performance scores for Low, Medium, and High clusters were determined to be
0.68, 0.43 and 0.62 respectively, with precision of 0.77, 0.49 and 0.66 when evaluated with
the test set (consisting of 4858 samples). Lower performance with the Medium cluster
reflects the inherent difficulty of classifying transitional states in natural systems. The
consistent level of accuracy across test and cross-validation demonstrates the generalised
performance of the RF classifier, while not accurate, was deemed sufficiently informative
for the purpose of assessing the influence of reef characteristics.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual predictors were systematically (one-at-a-time) varied across a grid of
50 equally spaced points spanning the predictor's range, while all other features were left
at their modelled values. The procedure creates (1) a modified test dataset, where the
target predictor was set to the grid value for all samples, (2) probabilistic predictions
for all three cluster classes using the trained RF classifier, and (3) the mean predicted
probability for each class across all test observations. Use of the median predicted
probability was also assessed but was found not to substantially influence the
interpretation of results (and so, are not shown here). The process isolates the marginal
effect of each predictor by averaging over the natural distribution of all other
representative reef characteristics in the test dataset. Additionally, to assess possible
interaction between reef depth and incoming connectivity, two-way partial dependence
analysis was performed on grids along the range of these two features.
